---
title: "amazonia_crop"
format: html
editor_options: 
  chunk_output_type: inline
---

# Load and Crop Amazonia Basin Data

This document loads raw dam and river data, crops to the Amazonia basin (South America Level 2), builds a directed river network, snaps dams to the network, and visualizes. Then saves as RDS

**Note:** `dam_id` == `grand_id`
---

# Load Libraries

```{r}
#| output: false 
library(sfnetworks)   
library(sf)          
library(tidygraph)
library(dplyr)       
library(tidyverse)   
library(ggplot2)     
library(janitor)     
```

---

# Load Raw Data

Load global dam databases and South America hydrography. Paths assume data in `capstone/netzerohydro/data/raw/`.

```{r}
# Current dams (GRanD)
current_dams <- read_sf("../../../../../capstone/netzerohydro/data/raw/GRanD_Version_1_3") %>% 
  janitor::clean_names()

# Future hydropower dams (FHReD)
future_dams <- read_csv("../../../../../capstone/netzerohydro/data/raw/FHReD_2015_future_dams/FHReD_2015_future_dams_Zarfl_et_al_beta_version.csv") %>% 
  janitor::clean_names()

# HydroBASINS
basin_sa <- read_sf("../../../../../capstone/netzerohydro/data/raw/hybas_sa_lev01-12_v1c/hybas_sa_lev02_v1c.shp") %>% 
  janitor::clean_names()

# World rivers + FFR (combined RDS)
world_rivers_ffr <- readRDS("../../../../../capstone/netzerohydro/data/cleaned_v1/world_rivers_ffr.rds")
```

---

# Make Future Dams Spatial and Crop to Amazonia

```{r}
# Make future dams df spatial 
future_dams <- st_as_sf(future_dams,
  coords = c("lon_cleaned", "lat_cleaned"), 
  crs = st_crs(current_dams))

# Filter water basin to Amazonia (Level 2)
amazon_basin <- basin_sa %>%
  filter(hybas_id == 6020006540)

# Filter rivers to order 3 
sa_rivers_3up <- world_rivers_ffr %>% 
  filter(ord_stra >= 3,
         continent == "South America")

sa_rivers_5up <- world_rivers_ffr %>% 
  filter(ord_stra >= 5,
         continent == "South America")

# Ensure geometries are valid for spatial operations
current_dams <- st_make_valid(current_dams)
future_dams <- st_make_valid(future_dams)
amazon_basin <- st_make_valid(amazon_basin)

# Crop dams and rivers to Amazonia polygon
amazon_current_dams <- st_intersection(amazon_basin, current_dams)
amazon_future_dams <- st_intersection(amazon_basin, future_dams)

amazon_rivers_3up <- sa_rivers_3up %>%
  st_filter(y = amazon_basin, .predicate = st_intersects)

amazon_rivers_5up <- sa_rivers_5up %>%
  st_filter(y = amazon_basin, .predicate = st_intersects)
```

---

# Standardize to dam_id and Check for Duplicates

```{r}
# Rename grand_id to dam_id (GRanD)
amazon_current_dams <- amazon_current_dams %>% rename(dam_id = grand_id)

# Convert to character for consistent joins
amazon_current_dams <- amazon_current_dams %>% mutate(dam_id = as.character(dam_id))
amazon_future_dams <- amazon_future_dams %>% mutate(dam_id = as.character(dam_id))

# Sanity check: no dam should appear in both current and future
any(amazon_current_dams$dam_id %in% amazon_future_dams$dam_id)
```

---

# Convert Rivers to LINESTRING

`sfnetwork` requires LINESTRING geometry. HydroRIVERS may store MULTILINESTRING; we cast to LINESTRING (splits multi-part lines into single segments).

```{r}
# Testing with order 3+
amazon_rivers <- st_cast(amazon_rivers_3up, "LINESTRING")
```

---

# Create River Network

Build a directed spatial network from river edges. Edge direction follows the flow path (first point to last point). `weight` = edge length in meters for distance calculations.

```{r}
net <- as_sfnetwork(amazon_rivers, directed = TRUE) %>%
  activate("edges") %>%
  mutate(weight = edge_length())
```

---

# Prepare Dams for Snapping

```{r}
# CRS must match for snapping
st_crs(amazon_current_dams) == st_crs(amazon_rivers)
st_crs(amazon_future_dams) == st_crs(amazon_rivers)

# GRanD may store dams as polygons; convert to point for blending
if (st_geometry_type(amazon_current_dams, by_geometry = FALSE) != "POINT") {
  amazon_current_dams <- st_point_on_surface(amazon_current_dams)
}

# Flag dam type for downstream filtering
amazon_current_dams <- amazon_current_dams %>% mutate(is_current_dam = TRUE)
amazon_future_dams <- amazon_future_dams %>% mutate(is_current_dam = FALSE)

# Store counts for snapping diagnostics
n_current_input <- nrow(amazon_current_dams)
n_future_input <- nrow(amazon_future_dams)
```

---

# Snap Dams to River Network

`st_network_blend()` snaps each dam point to the nearest edge, splits that edge, and inserts the dam as a node. Dams beyond `tolerance` are not snapped. Amazon scale uses 30 km tolerance (vs 1 km for Nepal).

```{r}
# Combine dams; only dam_id and is_current_dam needed for blending
all_dams <- rbind(
  amazon_current_dams %>% select(dam_id, is_current_dam),
  amazon_future_dams %>% select(dam_id, is_current_dam)
)

net_with_dams <- net %>% st_network_blend(all_dams, tolerance = 30000)

# Snapping summary
dam_nodes_snapped <- net_with_dams %>%
  activate("nodes") %>%
  as_tibble() %>%
  filter(!is.na(is_current_dam)) %>%
  pull(dam_id)

n_current_snapped <- sum(net_with_dams %>% activate("nodes") %>% as_tibble() %>% pull(is_current_dam) == TRUE, na.rm = TRUE)
n_future_snapped <- sum(net_with_dams %>% activate("nodes") %>% as_tibble() %>% pull(is_current_dam) == FALSE, na.rm = TRUE)

cat("=== Dam Snapping Summary ===\n")
cat("Current: Input", n_current_input, "| Snapped", n_current_snapped, "\n")
cat("Future: Input", n_future_input, "| Snapped", n_future_snapped, "\n")

# List dams that failed to snap (beyond tolerance or off-network)
current_dams_missing <- setdiff(amazon_current_dams$dam_id, dam_nodes_snapped)
future_dams_missing <- setdiff(amazon_future_dams$dam_id, dam_nodes_snapped)

if (length(current_dams_missing) > 0) cat("Current dams not snapped:", length(current_dams_missing), "\n")
if (length(future_dams_missing) > 0) cat("Future dams not snapped:", length(future_dams_missing), "\n")
```

---

# Extract Dam Nodes and Build Graph

Extract node IDs and sf geometries for current and future dams, then plot the river network with dams overlaid.

```{r}
# Node indices for connectivity matrices (if extending to connectivity_df workflow)
current_dam_nodes <- net_with_dams %>%
  activate("nodes") %>%
  mutate(node_id = row_number()) %>%
  as_tibble() %>%
  filter(is_current_dam == TRUE) %>%
  pull(node_id)

future_dam_nodes <- net_with_dams %>%
  activate("nodes") %>%
  mutate(node_id = row_number()) %>%
  as_tibble() %>%
  filter(is_current_dam == FALSE) %>%
  pull(node_id)

# sf objects for plotting
current_dam_nodes_sf <- net_with_dams %>%
  activate("nodes") %>%
  filter(is_current_dam == TRUE) %>%
  st_as_sf()

future_dam_nodes_sf <- net_with_dams %>%
  activate("nodes") %>%
  filter(is_current_dam == FALSE) %>%
  st_as_sf()

# Graph: river network (gray) with current dams (red) and future dams (blue)
ggplot() +
  geom_sf(data = st_as_sf(net_with_dams, "edges"), color = "gray40", size = 0.2) +
  geom_sf(data = current_dam_nodes_sf, color = "red", size = 0.5) +
  geom_sf(data = future_dam_nodes_sf, color = "blue", size = 0.5) +
  theme_minimal() +
  labs(
    title = "Amazonia: Current dams (red) and Future dams (blue) on river network",
    subtitle = paste0("Current: ", n_current_snapped, " | Future: ", n_future_snapped, " snapped")
  )
```
## Amazonia by Strahler River Order

```{r}
order_colors <- c(
  "3" = "yellow",
  "4" = "olivedrab",
  "5" = "darkgreen",
  "6" = "lightblue",
  "7" = "blue",
  "8" = "darkblue",
  "9" = "purple",
  "10" = "black"
)

ggplot() +
  geom_sf(data = amazon_basin, fill = NA, color = "black", linewidth = 0.4) +
  geom_sf(
    data = amazon_rivers_3up %>% mutate(ord_stra_chr = as.character(ord_stra)),
    aes(color = ord_stra_chr),
    linewidth = 0.25, alpha = 0.9
  ) +
  scale_color_manual(values = order_colors, name = "ord_stra") +
  theme_minimal() +
  labs(
    title = "Amazonia rivers colored by Strahler order (ord_stra)"
  )
```

---

# Save Cropped Data (Optional)

Save RDS files for use in downstream connectivity analysis (e.g., `connectivity_df`-style reach and cascade workflows).

```{r}
saveRDS(amazon_basin, "../../../../../capstone/netzerohydro/data/cleaned_v1/amazon_basin.rds")
saveRDS(amazon_current_dams, "../../../../../capstone/netzerohydro/data/cleaned_v1/amazon_current_dams.rds")
saveRDS(amazon_future_dams, "../../../../../capstone/netzerohydro/data/cleaned_v1/amazon_future_dams.rds")
saveRDS(amazon_rivers_3up, "../../../../../capstone/netzerohydro/data/cleaned_v1/amazon_rivers_3up.rds")
saveRDS(net_with_dams, "../../../../../capstone/netzerohydro/data/cleaned_v1/amazon_net_with_dams.rds")
```
