---
title: "model_attempt1"
format: html
---

# Connectivity Classification & Suitability Model (MCDA:AHP)

```{r}
#| output: false
library(tidyverse)
```

## Create fake data

```{r}
dam_df <- data.frame(
  dam_node = c("2001", "2002", "2003", "2004", "2005", "2006", "2007", "2008", "2009", "2010"), 
  trunk_id = c(1, 1, 1, 2, 2, 2, 3, 3, 4, 4), 
  upstream_dam_dist = c(NA, 10, 30, NA, 3, NA, 30, 200, NA, 2), 
  downstream_dam_dist = c(20, 10, NA, NA, 15, 30, 100, 10, NA, 14)
)

dam_df
```

## Model 1

### Find connectivity classification from distances

Define:

-   Cascade dam: Downstream & upstream dam within 15

-   FFR dam: No upstream, no downstream dam

-   Upstream dam: There is a downstream dam in 50km

-   Downstream dam: There is an upstream dam in 50km 

```{r}
# Connectivity classificantion column
dam_df <- dam_df %>% 
  mutate(connectivity_category = case_when(
    upstream_dam_dist <= 15 & downstream_dam_dist <= 15 ~ "cascade", 
    is.na(upstream_dam_dist) & is.na(downstream_dam_dist) ~ "FFR",
    is.na(upstream_dam_dist) | downstream_dam_dist <= 50 ~ "upstream", 
    upstream_dam_dist <= 50 | is.na(downstream_dam_dist) ~ "downstream", 
    TRUE ~ NA
  ))

# Column for casade only - not necessary 
dam_df <- dam_df %>% 
  mutate(cascade = ifelse(upstream_dam_dist <= 15 & downstream_dam_dist <= 15, 1, 0))

dam_df 
```

# Model 2

### Calculate Suitability Score from classification + CSI + population + PA

Link: <https://cran.r-project.org/web/packages/RMCDA/vignettes/RMCDA_guidelines.html>

**MCDA: Multi-Criteria Decision Analysis**

-   **Analytic Hierarchy Process (AHP)**: pairwise comparisons

    -   \<10 alternatives, informal weights, and a ranking goal

### Fake data

```{r}
# Add to the fake dataset 
dam_df <- dam_df %>% 
  mutate(CSI = c(0.40, 0.10, 0.22, 0.90, 0.12, 0.25, 0.81, 0.70, 0.95, 0.07), 
         population = c(0,0,0,0,1,1,0,1,1,1), 
         pa = c(1,0,0,1,1,0,0,0,1,1)
         )

dam_df
```

Connectivity cat –\> Numeric (with cascade weighted more) 

```{r}
# Need to make connectivity numeric 
# increase the gap between your connectivity numeric scores

#connectivity_scores <- c("cascade" = 4, "downstream" = 3, "upstream" = 2, "FFR" = 1)
#connectivity_scores <- c("cascade" = 100, "downstream" = 10, "upstream" = 3, "FFR" = 1) 
connectivity_scores <- c("cascade" = 10, "downstream" = 5, "upstream" = 2, "FFR" = 1)

dam_df$connectivity <- connectivity_scores[dam_df$connectivity_category]
```

### **Step 1: Make the df a `decision_matrix` (df –\> matrix)**

```{r}
library(RMCDA)

# Make df a matrix (RMCDA expects matrix)
decision_matrix <- as.matrix(dam_df[, c("connectivity", "CSI", "population", "pa")])
rownames(decision_matrix) <- dam_df$dam_node # Make the rows, the dam ids

```

### **Step 2: Define Criteria Weights via AHP**

What the matrix numbers mean:

+---------------+--------------------------------------------+
| Value         | Meaning                                    |
+===============+============================================+
| 1             | Equal importance                           |
+---------------+--------------------------------------------+
| 3             | Moderately more important                  |
+---------------+--------------------------------------------+
| 5             | Strongly more important                    |
+---------------+--------------------------------------------+
| 7             | Very strongly more important               |
+---------------+--------------------------------------------+
| 9             | Extremely more important                   |
+---------------+--------------------------------------------+
| 1/3, 1/5...   | The inverse — other criterion is preferred |
|               |                                            |
|               | The column is preferred over the row       |
+---------------+--------------------------------------------+

My thought:

-   connectivity \>\> CSI \>\> population = pa

(The diagonal is always 1 (criterion vs itself), and the lower triangle is always the reciprocal of the upper triangle).

```{r}
# Pairwise comparison matrix: comparing criteria on rows and columns
criteria_pairwise <- matrix(c(
  1,   3,   5,   5,   # connectivity vs all - connectivity is stronger than all
  1/3, 1,   3,   3,   # CSI vs all - CSI follows connectivity
  1/5, 1/3, 1,   1,   # population vs all - pop and pa are the same
  1/5, 1/3, 1,   1    # PA vs all
), nrow = 4, byrow = TRUE,
dimnames = list(
  c("connectivity", "CSI", "population", "pa"),
  c("connectivity", "CSI", "population", "pa")
))

# The numbers I used (3, 5) are reasonable starting points... but we can adjust them to whatever we think.

```

Another Pairwise comparison matrix of 1,5,7 
```{r}
# Pairwise comparison matrix: comparing criteria on rows and columns

# criteria_pairwise <- matrix(c(
#   1,   5,   7,   7,
#   1/5, 1,   3,   3,
#   1/7, 1/3, 1,   1,
#   1/7, 1/3, 1,   1
# ), nrow = 4, byrow = TRUE,
# dimnames = list(
#   c("connectivity", "CSI", "population", "pa"),
#   c("connectivity", "CSI", "population", "pa")
# ))
```

Need to make a pairwise comparisons of alternatives for each criteria for `apply.APA`

APA has 2 levels

-   *Level 1* — Criteria weights (what we already did) "How important is connectivity vs CSI vs population vs PA?" → Produces a weight for each criterion

    -   We have level 1 data because we already have numeric scores for each alternative on each criterion in the dataframe. Thius, we **don't need Level 2 AHP.**

    -   RMCDA's AHP wants alternative comparisons too, but because we will have real scores we don't need that — hence the dummy equal matrices just to extract the weights cleanly.

-   *Level 2* — Alternative comparisons per criterion "For this specific criterion, how does Site A compare to Site B, Site C, etc.?" → Done separately for each criterion → Produces a score for each alternative on each criterion

    -   Repeat this for every criterion — so with 4 criteria and 4 sites, that's **4 separate matrices** (why we make 4 matrices below)

```{r}
# Create a Dummy matrices (for the apply.APA) that has the same col and row as dam_df
n_alternatives <- nrow(decision_matrix)
equal_matrix <- matrix(1, nrow = n_alternatives, 
                       ncol = n_alternatives,
                       dimnames = list(rownames(decision_matrix), 
                                       rownames(decision_matrix)))

comparing.competitors <- list(
  connectivity = equal_matrix,
  CSI          = equal_matrix,
  population   = equal_matrix,
  PA           = equal_matrix
)
```

### **Step 3: Apply AHP –\> Get weights for each criteria!**

```{r}
# Create weights by applying the criteria_pairwise 
ahp_result <- apply.AHP(criteria_pairwise, comparing.competitors)
weights <- ahp_result[[1]]  # extracted criteria weights
print(weights)

# Extract just the 4 weights
w <- as.numeric(weights[[2]])
w
```

### **Step 4: Apply weights to decision matrix** 
Use the `topsis` packages to use the weights on the decision matrix

```{r}
library(topsis)

result <- topsis(decision_matrix, w, impacts = c("+", "-", "-", "-"))
print(result)

result$score
```

Add the scores to the dataframe with grades

```{r}
# Suitability score column
dam_df <- dam_df %>% 
  mutate(suitability_score = result$score)

# Give Letters to the scores
dam_df <- dam_df %>% 
  mutate(suitability_grade = case_when(
    suitability_score >= 0.75 ~ "A", 
    suitability_score >= 0.50 ~ "B",
    suitability_score >= 0.24 ~ "C",
    suitability_score >= 0 ~ "D"
  )
  )

dam_df
```

# DONT READ FURTHER 

**Step 3: Use weights in TOPSIS**

`TOPIS` reads the higher number a good

For population and pa, 0 = "good" and 1 = "bad".

For CIS, lower = "good" and higher = "bad".

Therefore, we need to flip those columns

```{r}
decision_matrix_flipped <- decision_matrix

#Flip cost criteria (CSI, population, PA) so lower = better becomes higher = better
decision_matrix_flipped[, "CSI"]        <- 1 - decision_matrix[, "CSI"]
decision_matrix_flipped[, "population"] <- 1 - decision_matrix[, "population"]
decision_matrix_flipped[, "pa"]         <- 1 - decision_matrix[, "pa"]
# Giving the binary criteria more leverage in the normalization without changing their relative importance
#decision_matrix
decision_matrix_flipped
```

To give scores to each dam

```{r}
topsis_result <- apply.TOPSIS(
  A = decision_matrix_flipped,
  w = w
)

print(topsis_result)

```

# Problem

2002 (cascade, low CSI, no pop, no pa) is scoring terribly

```{r}
# Check normalized matrix
norm <- apply(decision_matrix_flipped, 2, function(x) x / sqrt(sum(x^2)))
print(round(norm, 3))

# Check weighted normalized matrix
weighted_norm <- sweep(norm, 2, w, "*")
print(round(weighted_norm, 3))
```

Manually calculating scores

```{r}
# Step 1: normalized matrix
norm <- sweep(decision_matrix_flipped, 2, 
              sqrt(colSums(decision_matrix_flipped^2)), "/")

# Step 2: weighted normalized
weighted_norm <- sweep(norm, 2, w, "*")

# Step 3: ideal best and worst
ideal_best  <- apply(weighted_norm, 2, max)
ideal_worst <- apply(weighted_norm, 2, min)

# Step 4: distances
d_best  <- sqrt(rowSums((weighted_norm - ideal_best)^2))
d_worst <- sqrt(rowSums((weighted_norm - ideal_worst)^2))

# Step 5: score
score <- d_worst / (d_best + d_worst)

print(round(norm["2002",], 3))
print(round(weighted_norm["2002",], 3))
print(paste("ideal best:", round(ideal_best, 3)))
print(paste("2002 d_best:", round(d_best["2002"], 3)))
print(paste("2002 d_worst:", round(d_worst["2002"], 3)))
print(paste("2002 score:", round(score["2002"], 3)))
print(round(score, 3))
```

```{r}
print(topsis_result)
```

The manual TOPSIS gives 2002 a score of 0.572, But apply.TOPSIS gives 2002 = 0.236...

## DITCH TOPSIS
