---
title: "trunk_id_exploration"
format: html
editor_options:
  chunk_output_type: inline
---

# Trunk ID Exploration in the Amazonia Basin

This document explores **backbone identifications** (`bb_id`) we have been referring to as trunk id and related attributes for the Amazonia basin using RDS objects from the `amazonia_crop` workflow.

A backbone (`bb_id`) is the **single downstream path** a reach belongs to, from its headwater until it hits a terminus (ocean, lake, or where it joins a bigger river).

```{r libraries}
#| message: false
#| warning: false

library(sf)        
library(dplyr)     
library(ggplot2)   
library(tidyverse) 

# connectivity function libraries
library(sfnetworks)
library(igraph)
# also sf and dplyr
```


## Load Amazonia Objects

```{r}
# Load Basin and Rivers and Dams
amazon_basin <- readRDS("../../../../../capstone/netzerohydro/data/cleaned_v1/amazon_basin.rds")
amazon_rivers_3up <- readRDS("../../../../../capstone/netzerohydro/data/cleaned_v1/amazon_rivers_3up.rds")
amazon_current_dams <- readRDS("../../../../../capstone/netzerohydro/data/cleaned_v1/amazon_current_dams.rds")
amazon_future_dams <- readRDS("../../../../../capstone/netzerohydro/data/cleaned_v1/amazon_future_dams.rds")
amazon_amazon_net_with_dams <- readRDS("../../../../../capstone/netzerohydro/data/cleaned_v1/amazon_amazon_net_with_dams.rds")

# Load South America Level 3 sub-basins
basin_sa_lev3 <- read_sf(
  "../../../../../capstone/netzerohydro/data/raw/hybas_sa_lev01-12_v1c/hybas_sa_lev03_v1c.shp"
) %>%
  janitor::clean_names()

# Keep Level 3 basins that intersect Amazonia
amazon_subbasins_lev3 <- basin_sa_lev3 %>%
  st_make_valid() %>%
  st_filter(amazon_basin, .predicate = st_intersects)

# Load South America Level 4 sub-basins
basin_sa_lev4 <- read_sf(
  "../../../../../capstone/netzerohydro/data/raw/hybas_sa_lev01-12_v1c/hybas_sa_lev04_v1c.shp"
) %>%
  janitor::clean_names()

# Keep Level 4 basins that intersect Amazonia
amazon_subbasins_lev4 <- basin_sa_lev4 %>%
  st_make_valid() %>%
  st_filter(amazon_basin, .predicate = st_intersects)

```


## Amazonia Rivers Colored by Backbone ID (`bb_id`)

Each **backbone river** is a single, contiguous flow path from a headwater to a terminus (sink), identified by `bb_id`. Coloring by `bb_id` reveals the trunk structure of the network.

```{r}
# test <- amazon_rivers_3up %>%
#   mutate(
#     bb_id_chr    = as.character(bb_id),
#     bb_id_prefix = substr(bb_id_chr, 1, 2)   # first 4 digits
#   )

ggplot() +
  geom_sf(data = amazon_basin, fill = NA, color = "black", linewidth = 0.4) +
  geom_sf(data = amazon_rivers_3up,
          aes(color = as.factor(bb_id)),
          linewidth = 0.2, alpha = 0.9) +
  guides(color = "none") +  # hide legend; too many categories
  theme_minimal() +
  labs(
    title = "Amazonia rivers colored by backbone river ID (bb_id)",
    subtitle = "Each color represents one headwater-to-terminus backbone river"
  )
```

## Amazonia Rivers with Level 4 basins overlayed

```{r}
bbox_amazon <- st_bbox(amazon_basin)

ggplot() +
  geom_sf(data = amazon_basin, fill = NA, color = "black", linewidth = 0.4) +
  geom_sf(data = amazon_subbasins_lev4, fill = NA, color = "black", linewidth = 0.5) +
  geom_sf(data = amazon_rivers_3up,
          aes(color = as.factor(bb_id)),
          linewidth = 0.2, alpha = 0.9) +
  guides(color = "none") +
  coord_sf(
    xlim = c(bbox_amazon["xmin"], bbox_amazon["xmax"]),
    ylim = c(bbox_amazon["ymin"], bbox_amazon["ymax"]),
    expand = FALSE
  ) +
  theme_minimal() +
  labs(
    title = "Amazonia rivers colored by backbone river ID (bb_id)",
    subtitle = "Level 4 HydroBASINS sub-basins within the Amazon main basin"
  )
```
Get list of hybas basins with area we can choose an appropriate sized one. 
```{r}
amazon_subbasins_lev4 %>%
  st_drop_geometry() %>% 
  select(
    hybas_id,
    sub_area   = any_of("sub_area"),
    up_area    = any_of("up_area"),
    
  ) %>%
  arrange(desc(coalesce(sub_area, up_area))) %>%  # largest first
  head(50)
```

## Mapping all subbasins
```{r}
# List of sub-basins (or slice to top N to avoid too many panels)
subbasins_to_plot <- amazon_subbasins_lev4 %>%
  mutate(hybas_id_chr = as.character(hybas_id))

# For each sub-basin, get intersecting rivers and attach hybas_id
rivers_facet <- subbasins_to_plot %>%
  split(.$hybas_id) %>%
  lapply(function(sb) {
    rivers_one <- amazon_rivers_3up %>%
      st_filter(sb, .predicate = st_intersects)
    if (nrow(rivers_one) == 0) return(NULL)
    rivers_one$hybas_id_chr <- sb$hybas_id_chr[1]
    rivers_one
  }) %>%
  bind_rows()

# Same for polygon layer (one row per sub-basin for outline)
subbasins_facet <- subbasins_to_plot %>%
  filter(hybas_id_chr %in% unique(rivers_facet$hybas_id_chr))
```

```{r}
ggplot() +
  geom_sf(data = subbasins_facet, fill = NA, color = "black", linewidth = 0.4) +
  geom_sf(
    data = rivers_facet,
    aes(color = as.factor(bb_id)),
    linewidth = 0.35,
    alpha = 0.9
  ) +
  facet_wrap(~ hybas_id_chr) +   # drop scales = "free"
  guides(color = "none") +
  theme_minimal() +
  labs(
    title = "Amazonia Level-4 sub-basins: backbones (bb_id)",
    subtitle = "One panel per sub-basin"
  )
```
## Map a singular smaller basin to examine backbones

## Connectivity function example using this basin to see cascades
```{r}
# Load the connectivity function
# source(here::here("MEDS", "capstone", "connectivity", "connectivity_function.R"))
source(here::here("..", "connectivity", "connectivity_function.R"))

# ---- 1) Define the sub-basin and crop rivers ----
# One Level-4 HydroBASINS polygon; rivers restricted to this area for mapping and network.
one_subbasin <- amazon_subbasins_lev4 %>%
  filter(hybas_id == 6040345000)

# One amazon basin river network cropped to the selected sub-basin
rivers_one <- amazon_rivers_3up %>%
  st_filter(one_subbasin, .predicate = st_intersects) %>%
  mutate(
    bb_id_chr    = as.character(bb_id),
    bb_id_prefix = substr(bb_id_chr, 1, 4)   # for plotting trunks by first 4 digits
  )

# ---- 2) Build the directed river network for this sub-basin ----
# as_sfnetwork(..., directed = TRUE) gives flow direction; weight = edge length (m) for distances.
# The directed network is needed for the function
net_one <- as_sfnetwork(rivers_one, directed = TRUE) %>%
  activate("edges") %>%
  mutate(weight = edge_length())

# ---- 3) Crop dams to the sub-basin and flag current vs future ----
# The function expects nodes with dam_id and is_current_dam (TRUE/FALSE). We keep only geometry + those for blending.
dams_current_one <- dams_current_one %>%
  mutate(is_current_dam = TRUE) %>%
  select(dam_id, is_current_dam, geometry)

dams_future_one <- dams_future_one %>%
  mutate(is_current_dam = FALSE) %>%
  select(dam_id, is_current_dam, geometry)

all_dams_one <- rbind(dams_current_one, dams_future_one)

# ---- 4) Snap dams onto the network (blend) ----
# st_network_blend inserts each dam as a node on the nearest edge within tolerance (m). Edges keep bb_id from rivers_one.
net_with_dams_one <- net_one %>%
  st_network_blend(all_dams_one, tolerance = 30000)

# ---- 5) Run connectivity on the sub-basin network ----
# No subbasin/hybas_id args: the network already only contains this basin. same_trunk_only = TRUE restricts upstream/downstream to current dams on the same bb_id; cascade_threshold_km limits cascade to within that distance (km).
reach_df_one <- connectivity_from_network(
  net_with_dams_one,
  cascade_threshold_km = 500,
  same_trunk_only      = TRUE
)
print(reach_df_one)
```

```{r}
library(ggnewscale)

# Which future dams are in cascade (from sub basin connectivity matrix)
cascade_ids <- reach_df_one %>%
  filter(cascade_status == "cascade") %>%
  pull(dam_id)

dams_future_cascade <- dams_future_one %>% filter(dam_id %in% cascade_ids)
dams_future_other   <- dams_future_one %>% filter(!dam_id %in% cascade_ids)

# One color per cascade dam (for identification)
n_cascade <- nrow(dams_future_cascade)
cascade_pal <- setNames(
  scales::hue_pal()(max(1, n_cascade)),
  sort(unique(dams_future_cascade$dam_id))
)

ggplot() +
  # Sub-basin outline
  geom_sf(data = one_subbasin, fill = NA, color = "black", linewidth = 0.4) +
  # Rivers: trunk colors (first color scale)
  geom_sf(data = rivers_one, aes(color = as.factor(bb_id_prefix)), linewidth = 0.5) +
  scale_color_discrete(name = "bb_id (first 4 digits)") +
  # Second color scale for dams
  ggnewscale::new_scale_color() +
  geom_sf(data = dams_current_one, color = "red", size = 2) +
  geom_sf(data = dams_future_other, color = "gray60", size = 2) +
  geom_sf(data = dams_future_cascade, aes(color = as.factor(dam_id)), size = 2.5) +
  scale_color_manual(
    name = "Future dams (cascade)",
    values = cascade_pal,
    guide = guide_legend(override.aes = list(size = 3))
  ) +
  theme_minimal() +
  labs(
    title = "Sub-basin: trunks by bb_id; future cascade dams by dam_id",
    subtitle = "Red = current dams; gray = future not in cascade"
  )

```
### Interactive with TMAP

```{r}
#| message: false
#| warning: false
library(tmap)
tmap_mode("view")   # interactive mode

tm_shape(one_subbasin) +
  tm_borders(col = "black", lwd = 2) +
tm_shape(rivers_one) +
  tm_lines(col = "bb_id_prefix", palette = "Set2", title.col = "bb_id (first 4 digits)") +
tm_shape(dams_current_one) +
  tm_symbols(size = 0.3, col = "red", id = "dam_id") +
tm_shape(dams_future_other) +
  tm_symbols(size = 0.3, col = "gray60", id = "dam_id") +
tm_shape(dams_future_cascade) +
  tm_symbols(size = 0.35, col = "dam_id", id = "dam_id", palette = "Set1")
```

## Interactive with Leaflet

```{r}
library(mapview)

# Optional: set colors to match your ggplot (one color per cascade dam, etc.)
mapview(one_subbasin, color = "black", fill = NA, lwd = 2) +
  mapview(rivers_one, zcol = "bb_id_prefix", legend = TRUE) +
  mapview(dams_current_one, color = "red", col.regions = "red", label = dams_current_one$dam_id, cex = 3) +
  mapview(dams_future_other, color = "gray60", col.regions = "gray60", label = dams_future_other$dam_id, cex = 3) +
  mapview(dams_future_cascade, zcol = "dam_id", label = dams_future_cascade$dam_id, cex = 3) # Cex is size
```


## One subbasin with rivers
```{r}
ggplot() +
  geom_sf(data = one_subbasin, fill = NA, color = "black", linewidth = 0.4) +
  geom_sf(data = rivers_one, aes(color = as.factor(bb_id)), linewidth = 0.5) +
  #guides(color = "none") +
  theme_minimal() +
  labs(
    title = "Manually selected Level-4 sub-basin",
    color = "bb_id"
  )
```

Amazon_rivers_3up has one row per river reach (segment). Many reaches share the same bb_id because they belong to the same backbone/trunk.
bb_summary collapses those many reach rows into one row per bb_id, so you can rank/plot “major trunks” without duplicating backbone attributes across thousands of reaches.

```{r}
bb_summary <- amazon_rivers_3up %>%
  st_drop_geometry() %>%          # drop geometry so summarise is faster (attributes only)
  group_by(bb_id) %>%             # group all reaches that share the same backbone ID
  summarise(
    bb_name    = first(bb_name),  # backbone name (same for all reaches in this bb_id)
    bb_len_km  = first(bb_len_km),# total backbone length (precomputed backbone attribute)
    bb_dis_ord = first(bb_dis_ord),# discharge-based order for the backbone (precomputed)
    bb_vol_tcm = first(bb_vol_tcm),# total backbone volume (precomputed)
    .groups = "drop"              # return an ungrouped data frame
  )

top_n <- 5
```

## Largest Rivers by length (could switch to volume)
```{r}
top_bb_len <- bb_summary %>%
  arrange(desc(bb_len_km)) %>%
  slice_head(n = top_n) %>%
  pull(bb_id)

rivers_len <- amazon_rivers_3up %>%
  mutate(trunk_group = ifelse(bb_id %in% top_bb_len, as.character(bb_id), "Other"))

ggplot() +
  geom_sf(data = amazon_basin, fill = NA, color = "black", linewidth = 0.4) +
  geom_sf(data = rivers_len, aes(color = trunk_group), linewidth = 0.3, alpha = 0.95) +
  scale_color_manual(
    values = c(setNames(rep("grey75", 1), "Other"),
               setNames(RColorBrewer::brewer.pal(top_n, "Set1"), as.character(top_bb_len))),
    name = paste0("Top ", top_n, " bb_id by bb_len_km")
  ) +
  theme_minimal() +
  labs(title = "Major trunks highlighted by backbone length (bb_len_km)")
```


```{r}
library(patchwork)

p_reach <- ggplot() +
  geom_sf(data = one_subbasin, fill = NA, color = "black", linewidth = 0.4) +
  geom_sf(data = rivers_one, aes(color = as.factor(hyriv_id)), linewidth = 0.5) +
  guides(color = "none") +
  scale_color_discrete(name = "Reach (HYRIV_ID)") +
  theme_minimal() +
  labs(title = "By reach (hyriv_id)")

p_bb <- ggplot() +
  geom_sf(data = one_subbasin, fill = NA, color = "black", linewidth = 0.4) +
  geom_sf(data = rivers_one, aes(color = as.factor(bb_id_prefix)), linewidth = 0.5) +
  guides(color = "none") +
  scale_color_discrete(name = "bb_id (first 4 digits)") +
  theme_minimal() +
  labs(title = "By backbone (bb_id)")

p_reach + p_bb
```




## Backbone Attributes: Definitions (from HydroROUT / HydroATLAS)

For each backbone (`bb_id`), HydroROUT / HydroATLAS provides:

- **`bb_id`**: Backbone River Identifier. Represents the contiguous river unit (flow path from source/headwater to sink/terminus).
- **`bb_name`**: Backbone river name.
- **`bb_len_km`**: Backbone river length. Sum of `length_km` across all reaches belonging to the same `bb_id`.
- **`bb_dis_ord`**: Backbone river discharge order. River order (`riv_ord`) of the most downstream reach of that backbone (discharge-based order).
- **`bb_vol_tcm`**: Backbone river volume. Sum of `volume_tcm` across all reaches in the backbone.
- **`bb_ocean`**: Ocean connectivity flag. Equals 1 if the backbone is directly connected to the ocean (its terminus drains to the ocean), 0 otherwise.

At the **reach** level we also have:

- **`riv_ord`**: River order (based on long-term average discharge). Larger values typically indicate more downstream, higher-discharge reaches.


## Takeaways

- **`bb_id`** partitions the Amazonia river network into coherent trunk units (from headwaters to termini).
- **`bb_len_km`** and **`bb_vol_tcm`** indicate the physical scale of each backbone; longest and largest-volume trunks correspond to major rivers.
- **`bb_ocean`** distinguishes backbones that ultimately drain to the ocean from those that terminate internally (e.g., in lakes or endorheic areas).
- **`bb_dis_ord`** summarizes each backbone's importance in terms of downstream discharge, complementing reach-level `riv_ord`.

These backbone attributes are useful for:

- Restricting **cascade** definitions to dams on the same trunk (`bb_id`).
- Prioritizing **major trunks** (high `bb_dis_ord`, `bb_len_km`, `bb_vol_tcm`) for conservation or impact assessment.

